---
title: "Lab 01: Performance Comparison in R"
author: ""
date: "`r Sys.Date()`"
format: html
---

## Introduction

In this lab, you will implement a Monte Carlo simulator in **R** and compare its performance to the Python versions you explored in the tutorial. This exercise will help you analyze the efficiency of different programming languages in financial simulations.

## Learning Outcomes

- Implement a Monte Carlo simulation in **R**
- Measure execution time and compare with Python
- Discuss performance differences between R and Python

## Task 1: Implement a Monte Carlo Simulator in R

Using the same **Black-Scholes-Merton** model from the tutorial, write an R function that simulates asset price paths.

### **Starter Code**
The following R function is a template. Complete the missing parts to implement the Monte Carlo simulation:

```{r}
monte_carlo_simulation_r <- function(M, I, S0=36, T=1, r=0.06, sigma=0.2) {
  dt <- T / M
  S <- matrix(0, nrow = M+1, ncol = I)
  S[1, ] <- S0
  rn <- matrix(rnorm(M * I), nrow = M, ncol = I)
  
  for (t in 2:(M+1)) {
    S[t, ] <- S[t-1, ] * exp((r - 0.5 * sigma^2) * dt + sigma * sqrt(dt) * rn[t-1, ])
  }
  return(S)
}
```

### **Run the Simulation**
- Set `M = 10000` (number of time steps)
- Set `I = 10000` (number of paths)
- Measure execution time using `system.time()`

```{r}
M <- 10000
I <- 10000

start_time <- Sys.time()
S_r <- monte_carlo_simulation_r(M, I)
end_time <- Sys.time()

execution_time_r <- end_time - start_time
print(paste("Elapsed time in seconds:", execution_time_r))
```

## Task 2: Compare Performance with Python

1. Compare your **R execution time** with the **Python implementations** from the tutorial.
2. Write a short paragraph discussing:
   - Is R faster or slower than Python for this simulation?
   - What factors might explain the difference in execution time?
   - Could optimization techniques (e.g., vectorization, parallel computing) improve R's performance?

## Submission Instructions

- Submit your completed R script (`lab01-performance-comparison.qmd`).
- Include your analysis as a short summary at the end of the document.
